---
title: 'Using *sabinaHSBM* for link prediction and network reconstruction with "conditional_missing" method'
output:
  pdf_document: 
    latex_engine: xelatex
---

> [!NOTE]  
> Highlights information that users should take into account, even when skimming.

This guide shows a simple use case based on the "conditional_missing" method in *sabinaHSBM* to identify missing links (false negatives) in an incomplete network. 

We use the example dataset `dat` included in the package to show key functionalities, including data preparation, link prediction, and network reconstruction.


# Requirements and installation

There are **two ways** to use the package:

## Option A: Docker (recommended)

**Non-native UNIX users** or those who prefer to avoid manual dependency installation can use the provided ready-to-use Docker image which includes R, Python, all dependencies, and *sabinaHSBM* pre-installed (see Supporting Information S1 for Docker container setup).

Once the container is running, simply load the package in your R session with:

```{r docker, message=FALSE, warning=FALSE, eval = FALSE}
library(sabinaHSBM)
```

## Option B: Native UNIX installation

**Native UNIX users** can run *sabinaHSBM* locally **if** their system includes:

   - R (version $\leq$ 4.0.4) with all required R packages (listed below)
   - Python $\leq$ with the `graph-tool` library (version $\leq$ 2.45)

```{r setup, message=FALSE, warning=FALSE, eval = FALSE}
install.packages("remotes")
# If the package is not installed, install sabinaHSBM from GitHub
 if (!requireNamespace("sabinaHSBM", quietly = TRUE)) {
   remotes::install_github("h-lima/sabinaHSBM")
 }

# Load sabinaHSBM
library(sabinaHSBM)
```

Install required R packages if not already available:

```{r install_depend, message=FALSE, warning=FALSE, eval = FALSE}
list.of.packages <- c(
  "dplyr",
  "reshape2",
  "reticulate",
  "stringr",
  "tidyr",
  "ROCR"
)

new.packages <-
    list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(new.packages) > 0) {
  install.packages(new.packages, dependencies = TRUE)
}

for (package in list.of.packages) {
  library(package, character.only = TRUE)
}
```

```{r load_env, include = FALSE}
library(dplyr)
library(kableExtra)
devtools::load_all("../")
```

```{r start_time}
# Record starting time (optional)
start_time <- Sys.time()
```

# Load example data

The dataset `dat` is a binary matrix representing a hypothetical species 
interactions network. Columns and rows correspond to two different types of 
nodes (e.g. hosts and parasites), and links (values of `1`, in red) represent interaction 
between them while `0` (in black) represent lack of an observed interaction. 
The network contains gaps, indicating potential missing links.

```{r load-data}
# Load the dataset
#data(dat, package = "sabinaHSBM")
load("../testing/dat.RData")
```

```{r plot-observed-matrix, echo=TRUE}
# Plot the simulated matrix
plot_interaction_matrix(dat)
```


# Preparing Input data for HSBM

The `hsbm.input` function cleans the dataset, and creates cross-validation folds by 
temporarily holding out subsets of observed links. Internally, it ensures that each 
fold masks a representative sample. In our example, it randomly splits the `r sum(dat)` 
observed links into five similar-sized subsets (because here we use 
5-fold cross-validation), holding out one subset per fold to test the
model’s ability to recover missing links.

```{r prepare-input}
# Prepare input object
myInput <- hsbm.input(
  dat,            # Binary matrix of observed links
  n_folds = 10     # Number of folds for cross-validation
)
```

```{r summary-input}
# Summarizes network characteristics
summary(myInput)
```


# Predicting missing links with "conditional_missing"

The `hsbm.predict` works directly with the processed input created by `hsbm.input`. 
It applies HSBM to each fold’s training network and computes posterior probabilities for links
and group assignments for nodes. The `conditional_missing` method focuses 
exclusively on predicting the conditional posterior probability of unobserved links (`0s`) 
given the network’s learned block structure, useful to identify missing links 
(i.e., unobserved host-parasite interactions likely to exist) in partially incomplete
networks—without casting doubt on any documented interactions.
Computation can be intensive, but parallelization across folds is supported with the `n_cores`
argument. 

```{r sequential-prediction}
# Predict missing links using HSBM
myPred <- hsbm.predict(
  myInput,          # Input data processed by hsbm.input()
  iter = 10000,      # Number of iterations 
  wait = 1000,      # Number of iterations for MCMC equilibration
  rnd_seed = 123,      # Sets seed in python environment for reproducibility
  method = "conditional_missing",  # Prediction method
  save_blocks = TRUE,  # Save group assignments 
  save_pickle = FALSE, # Save results as pickle files,
  save_plots = FALSE,  # Save hierarchical edge bundling plots
  n_cores = 2 # Number of cores to use
)
```

```{r save-pred, include = FALSE}
save("myPred", file = "mypred_s2.RData")
```

Predicted probabilities and group assignments are stored for each fold. Below, we extract 
the probabilities (*p*) and groups for fold 1. The lower the *p*, the less likely the link 
exists and the lower its priority; conversely, higher *p* values are the most promising 
missing interactions. 

```{r prediction-probs}
# View probabilities for fold 1
probabilities_fold1 <- myPred$probs[[1]]
head(probabilities_fold1)
```

The group assignments provide the hierarchical clustering structure of nodes for each fold. Let’s extract and examine the group assignments for fold 1:

```{r prediction-groups}
# View the group/block assignments for fold 1
groups_fold1 <- myPred$groups[[1]]

# Filter one type of nodes (nodes in columns)
vnames <- colnames(myPred$data)
groups_fold1 <- groups_fold1 %>% filter(name %in% vnames)
g_cols <- grep("^G", names(groups_fold1))
groups_fold1[g_cols] <- lapply(groups_fold1[g_cols], sort)

#print(head(groups_fold1))
print(lapply(myPred$groups,
             function(x){
                 g_cols <- grep("^G", names(x))
                 return(apply(x[g_cols], 2, function(y) length(unique(y))))
             }))
```

Hierarchical group assignments provide insight into how nodes (e.g., hosts) are organized 
across multiple levels. At the first level (G1), nodes are divided into specific 
groups, reflecting fine-scale patterns. Moving to higher levels (G2, G3, G4), 
these groups are progressively aggregated, revealing broader patterns and 
relationships or communities.

# Network Reconstruction

The `hsbm.reconstructed` function generates a reconstructed binary interaction matrix
by combining the probabilities across all folds and transforming them into a 
new binary (`0`/`1`) matrix/network from a user-specified threshold.

```{r network-reconstruction}
# Network reconstruction
myReconst <- hsbm.reconstructed(
  myPred,                # Predictions processed by hsbm.predict
  rm_documented = TRUE,  # Use of documented entries during validation
  threshold = "prc_closest_topright", # Binarization threshold
  consistency_matrix = "average_thresholded" # Combine fold predictions
)
```

This output includes the averaged probability matrix, the final reconstructed matrix, 
and evaluation metrics. These results highlight the predicted interactions, showcasing 
the method's ability to detect missing links effectively. Let’s explore some of the key 
outputs in detail.

```{r summary-reconstruction}
# View the reconstructed network summary and evaluation metrics
summary(myReconst)
```

```{r network-reconstruction2}
# Network reconstruction
myReconst2 <- hsbm.reconstructed(
  myPred,                # Predictions processed by hsbm.predict
  rm_documented = TRUE,  # Use of documented entries during validation
  consistency_matrix = "average_thresholded" # Combine fold predictions
)
```

```{r summary-reconstruction2}
# View the reconstructed network summary and evaluation metrics
summary(myReconst2)
```

The summary provides the number of missing links, as well as key evaluation metrics, 
such as the recovery link fraction (RLF).

```{r top-links}
# View the top potential missing links
top_links_df <- top_links(myReconst, 
                          n = 10, 
                          edge_type = "undocumented") # Type of edge to rank
print(top_links_df)
```

The `top-links` function identifies the undocumented links most likely to be missing 
links in a network predicted by HSBM.

```{r plot-reconstruction}
# View the reconstructed binary matrix
plot_interaction_matrix(myReconst$new_mat)
plot_interaction_matrix(myReconst2$new_mat)
```

```{r plot_reconstuction_missing, echo=TRUE}
# Plot the matrix with spurious
cols <- c("white", "red4")
dat_spur <- myReconst$data
new_missing <- which(myReconst$data == 0 & myReconst$new_mat == 1)
if(length(new_missing)!=0){
    dat_spur[new_missing] <- 2
    cols <- c(cols, "green4")
}
plot_interaction_matrix(dat_spur,
                        col = cols, order_mat = TRUE)
```

This document demonstrates the use of the *sabinaHSBM* package for network reconstruction. 
By applying the `conditional_missing` method, we showcased how missing links can be effectively 
identified and addressed in incomplete networks.

# Computing characteristics


```{r end_time}
# Show processing time and computer characteristics (optional)
end_time <- Sys.time()
cat("The processing time of this script took: ",
    round(as.numeric(difftime(end_time, start_time, units = "mins", 1))), "minutes\n")
```

This analysis was performed on a Dell Inspiron notebook with the following
characteristics:

- Processor (CPU): Intel Core i7-7500U @ 3.5GHz 
- Memory (RAM): 16 GB
- Operating System: Ubuntu 24.04 (run on a docker image)
- R Version: 4.3.3

