---
title: "Using sabinaHSBM for link prediction and network reconstruction using Hierarchical Stochastic Block Models"
subtitle: "Example on how to use it to identify missing and spurious links with the full_reconstruction method (with parallelized computation)"
output:
  word_document: default
  pdf_document: default
  html_document:
    toc: true
    toc_float: true
---


The *sabinaHSBM* package provides tools for link prediction and network reconstruction using hierarchical stochastic block models (HSBM). This document demonstrates a simple use for the `full_reconstruction` method, showcasing how it handles missing and spurious links in an incomplete and error-prone network. To enhance computational efficiency during the prediction links step we leverage parallelized computation.

We use the example dataset `dat` included in the package to show key functionalities, including data preparation, link prediction, and network reconstruction. 


# Loading Required Libraries

We begin by loading the required R packages. Missing packages are automatically installed.

```{r setup, message=FALSE, warning=FALSE}
# If the package is not installed, install it from GitHub
# if (!requireNamespace("sabinaHSBM", quietly = TRUE)) {
#   library(remotes)
#   remotes::install_github("h-lima/sabinaHSBM")
# }

# Load the sabinaHSBM package
#library(sabinaHSBM)
devtools::load_all("/home/project_hsbm/graph_tool/sabinaHSBM/")
```

If the required packages are not already installed, install them from CRAN

```{r install_depend, message=FALSE, warning=FALSE}
list.of.packages <- c(
  "dplyr",
  "parallel",
  "reshape2",
  "reticulate",
  "stringr",
  "tidyr",
  "ROCR",
  "data.table"
)

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(new.packages) > 0) {
  install.packages(new.packages, dependencies = TRUE)
}

for (package in list.of.packages) {
  library(package, character.only = TRUE)
}
```

```{r start_time}
# Record starting time (optional)
start_time <- Sys.time()
```

# Load data

The dataset `dat2` is a binary bipartite matrix representing a hypothetical species interactions network. Columns and rows correspond to two different types of nodes (e.g., hosts and parasites), and links (values of `1`, in red) represent interaction between them while `0` (in black) represent lack of observed interaction. The distribution of these links follows a structured pattern.

```{r load-data}
# Load the dataset
data(dat2, package = "sabinaHSBM")
dat<- dat2
```

```{r plot-data}
# Plot the original matrix
plot_interaction_matrix(adj_mat = dat, order_mat = FALSE)
```

To simulate spurious links and address the `full_reconstruction` method, we randomly add a small number of false positives (`1`) to the matrix. This controlled modification allows us to demonstrate how the method identifies both spurious links (observed but potentially erroneous interactions, false positives) and missing links (unobserved but likely interactions, false negatives).

```{r add-spurious-links, echo=TRUE}
# Add spurious links artificially
set.seed(123)
num_spurious <- ceiling(sum(dat) * 0.01) # Proportion of spurious links to add
spurious_links <- matrix(nrow = num_spurious, ncol = 2)

for (i in 1:num_spurious) {
   repeat {
     random_row <- sample(1:nrow(dat), 1)
     random_col <- sample(1:ncol(dat), 1)
     
     if (dat[random_row, random_col] == 0) {
       dat[random_row, random_col] <- 1
       spurious_links[i,] <- c(random_row, random_col)
       break
     }
   }
 }
```

```{r plot-dat_spur, echo=TRUE}
# Plot the matrix with spurious
plot_interaction_matrix(dat, order_mat = FALSE)
```


# Preparing Input data for HSBM

The `hsbm.input` function pre-processes the dataset, creating cross-validation folds and edge lists required for modeling. Here, we use 5-fold cross-validation and the `full_reconstruction` method.

```{r prepare-input}
# Prepare input data
n_folds <- 5       # Number of folds for cross-validation

myInput <- hsbm.input(
  dat,              # Binary bipartite matrix of observed links
  n_folds = n_folds,
  add_spurious = FALSE # Simulate spurious links
)
```

Only for this example, the following code assigns the "spurious_edge" label to the previously added spurious links. The purpose is to track these links throughout the analysis and network reconstruction process.

```{r add-spur}
myInput$edgelists <- lapply(myInput$edgelists,
                             function(x){
                                 x$x <- 1
                                 return(x)
                             })

for(i in 1:length(myInput$edgelists)){
    el <- myInput$edgelists[[i]]
    rows <- as.numeric(el[, 1]) + 1
    cols <- as.numeric(el[, 2]) - nrow(myInput$data) + 1
    rows_cols <- paste0(rows, "_", cols)
    spurious_rows_cols <- paste0(spurious_links[, 1], "_", spurious_links[, 2])
    spurious_loc <- which(rows_cols %in% spurious_rows_cols)
    el[spurious_loc, ]$x <- 1
    el[spurious_loc, ]$edge_type <- "spurious_edge"
    myInput$edgelists[[i]] <- el
}
```


```{r summary-input}
# Summarizes network characteristics
summary(myInput)
```


# Predict link probabilities

The `hsbm.predict` function applies the HSBM to predict probabilities of all links (observed and unobserved links). This step is crucial to identify spurious and missing links within the data, which are often present in incomplete or error-prone networks. The function works directly with the processed input created by `hsbm.input`.
This step can be computationally intensive when working with large datasets or numerous folds. To improve performance, we use parallelized computation, distributing tasks across multiple cores.  

```{r paarallel-prediction}
# Define the number of cores to use
#nCores <- detectCores() - 1
nCores <- 2

# Generate HSBM predictions
myPred <- hsbm.predict(
  myInput,           # Input data processed by hsbm.input()
  iter = 10000,      # Number of iterations
  wait= 10000,       # Number of iterations for MCMC equilibration
  method = "full_reconstruction", # Method for link prediction
  save_blocks = TRUE, 
  save_pickle = FALSE,
  n_cores = nCores
)
```

Predicted link probabilities and group assignments are stored for each fold. Below, we extract the link probabilities (p) and groups assignments of links for fold 1.

```{r prediction-probs}
# View probabilities for fold 1
probabilities_fold1 <- myPred$probs[[1]]
head(probabilities_fold1)
```

The group assignments provide the hierarchical clustering structure of nodes for each fold. Let’s extract and examine the group assignments for fold 1:

```{r prediction-groups}
# View the group/block assignments for fold 1
groups_fold1 <- myPred$groups[[1]]

# Filter one type of nodes (e.g., hosts, in columns)
vnames <- colnames(myPred$data)
groups_fold1 <- groups_fold1 %>% filter(names %in% vnames)

g_cols <- grep("^G", names(groups_fold1))
groups_fold1[g_cols] <- lapply(groups_fold1[g_cols], sort)

print(head(groups_fold1))
```

Hierarchical group assignments provide insight into how nodes (e.g., hosts) are organized across multiple levels. At the first level (G1), nodes are divided into specific groups, reflecting fine-scale patterns. Moving to higher levels (G2, G3, G4), these groups (or communities) are progressively aggregated, revealing broader patterns and relationships or communities.

Let's now look at the links for each fold filtered by type ("documented", "held out", and "spurious_edge") with a probability lower than 0.99.

```{r view-spur}
for(i in 1:5){
    cat("Fold ", i, "\n")
    cat("\tDocumented\n\t")
    print(dplyr::filter(myPred$probs[[i]], edge_type == "documented", p < 0.95))
    cat("\tHeld out\n\t")
    print(dplyr::filter(myPred$probs[[i]], edge_type == "held_out", p < 0.95))
    cat("\tSpurious edge\n\t")
    print(dplyr::filter(myPred$probs[[i]], edge_type == "spurious_edge", p < 0.95))
}
```

# Network Reconstruction

The `hsbm.reconstructed` function generates a reconstructed binary interaction matrix by combining predictions from all folds. The predicted matrix is binarized using a user-specified threshold.  

```{r network-reconstruction}
# Network reconstruction
myReconst <- hsbm.reconstructed(
  myPred,              # Predictions processed by hsbm.predict
  rm_documented=FALSE,
  threshold = 0.5,       # Binarization threshold
  new_matrix_method = "average_thresholded", # Combine fold predictions
  spurious_edges = TRUE
)
```

This output includes the final averaged probability matrix, the final reconstructed binary matrix, and evaluation metrics. These results highlight the predicted interactions, showcasing the method's ability to detect spurious and missing links effectively. Let’s explore some of the key outputs.

```{r plot-reconstruction}
# Visualize the reconstructed binary matrix
plot_interaction_matrix(myReconst$new_mat, order_mat = FALSE)
```

```{r summary-reconstruction}
# View the reconstruction summary and evaluation metrics
summary(myReconst)
```

The summary provides the number of spurious and missing links, as well as key evaluation metrics, such as the retained link recovery rate (RLRR).

Let's now examine the top 10 predicted links that are most likely to be spurious (false positives) by visualizing the probabilities of "documented" links.

```{r top-links}
# Visualize the top most likely spurious links
# top_links_spurious <- top_links(hsbm_out, n = 10, edge_type = "documented")
# print(top_links_spurious)
```


This document demonstrates the use of the *sabinaHSBM* package for network reconstruction. By applying the full reconstruction method, we showcased how spurious and missing links can be effectively identified and addressed in complex networks.


# Computing characteristics

Show processing time and computer characteristics

```{r end_time}
end_time <- Sys.time()
cat("The processing time of this script took: \n")
end_time - start_time
```

This analysis was performed on a Dynabook with the following characteristics:
  - Processor (CPU): Intel Core i7-1165G7 @ 2.80GHz (11th Gen)
  - Memory (RAM): 32 GB
  - Operating System: Windows 11 Pro, Version 24H2
  - R Version: 4.3.3
  

