---
title: 'Using *sabinaHSBM* for link prediction and network reconstruction with "marginal_all" method'
output:
  pdf_document:
    latex_engine: xelatex
---

```{r load_pkg, include = TRUE}
library(ggplot2)
library(RColorBrewer)
```

This guide shows a simple use case based on the "marginal_all" method in *sabinaHSBM* to 
detect both missing (false negatives) and/or spurious (false positives) links in an 
incomplete and/or error-prone network.

We use the artificially created dataset `dat2` included in the package to show key functionalities,
including data preparation, link prediction, and network reconstruction.

# Requirements and installation

There are **two ways** to use the package:

## Option A: Docker (recommended)

**Non-native UNIX users** or those who prefer to avoid manual dependency installation can 
use the provided **ready-to-use Docker image** (includes R, Python, all dependencies, and
*sabinaHSBM* pre-installed (see Supporting Information S1 for Docker container setup).

Once the container is running, simply load the package in your R session with:

```{r docker, message=FALSE, warning=FALSE, eval = FALSE}
library(sabinaHSBM)
```

## Option B: Native UNIX installation

**Native UNIX users** can run *sabinaHSBM* locally **if** their system includes:
   - R (version $\leq$ 4.0.4) with all required R packages (listed below)
   - Python $\leq$ with the `graph-tool` library (version $\leq$ 2.45)

```{r setup, message=FALSE, warning=FALSE, eval = FALSE}
install.packages("remotes")
# If the package is not installed, install sabinaHSBM from GitHub
 if (!requireNamespace("sabinaHSBM", quietly = TRUE)) {
   remotes::install_github("h-lima/sabinaHSBM")
 }

# Load sabinaHSBM
library(sabinaHSBM)
```

Install required R packages if not already available:

```{r install_depend, message=FALSE, warning=FALSE, eval = FALSE}
list.of.packages <- c(
  "dplyr",
  "reshape2",
  "reticulate",
  "stringr",
  "tidyr",
  "ROCR"
)

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(new.packages) > 0) {
  install.packages(new.packages, dependencies = TRUE)
}

for (package in list.of.packages) {
  library(package, character.only = TRUE)
}
```

```{r load_env, include = FALSE}
devtools::load_all("../")
```

```{r start_time}
# Record starting time (optional)
start_time <- Sys.time()
```


# Load example data

The dataset `dat2` is a binary matrix representing a hypothetical species interactions
network. Columns and rows correspond to two different types of nodes (e.g., hosts and parasites),
and links (values of `1`, in red) represent interactions between them while `0` (in black) 
represent lack of an observed interaction. These links are distributed in a pronounced 
structured pattern, to **focus the example on identifying spurious links**.

```{r load-data}
# Load the dataset
#data(dat2, package = "sabinaHSBM")
#dat <- dat2
load("../testing/dat2.RData")
```

```{r plot-data}
# Plot the original matrix
plot_interaction_matrix(adj_mat = dat, order_mat = FALSE)
```

To simulate spurious links and address the `marginal_all` method, we randomly add a small number
of false positives (`1`) to the matrix. This controlled modification allows us to demonstrate how
the method can work with both spurious links (observed but potentially erroneous interactions,
false positives) and missing links (unobserved but likely interactions, false negatives).


```{r add-spurious-links, echo=TRUE}
# Add spurious links artificially
set.seed(123)

n_spurious <- ceiling(sum(dat) * 0.02) 

# Find some random spurious
zero_indices <- which(dat == 0)
sampled_indices <- sample(zero_indices, n_spurious)
spurious_links <- arrayInd(sampled_indices, dim(dat))

total_spurious <- nrow(spurious_links)

dat[sampled_indices] <- 1
```

```{r plot-dat_spur, echo=TRUE}
# Plot the matrix with spurious
dat_spur <- dat
dat_spur[sampled_indices] <- 2
plot_interaction_matrix(dat_spur,
                        col = c("white", "red4", "blue1"),
                        order_mat = FALSE)
```


# Prepare input data for HSBM

The `hsbm.input` function cleans the dataset and creates cross-validation folds by 
temporarily holding out subsets of observed links. Internally, it ensures that each 
fold masks a representative sample. In our example, it randomly splits observed links 
into five similar-sized subsets (because here we use 5-fold cross-validation), holding out
one subset per fold to test the model’s ability to recover removed links. 
We set "max\_held\_per\_fold" to the number of spurious to keep the data more balanced, otherwise,
number of removed links would surpass the number of added spurious by a large margin 
hindering inference, since to detect spurious it is better to have the dataset as complete 
as possible.

```{r prepare-input}
# Prepare input data
n_folds <- 5     # Number of folds for cross-validation

myInput <- hsbm.input(
  dat,              # Binary matrix of observed links
  n_folds = n_folds,
  max_held_per_fold = total_spurious
)
```

For the purposes of this example, the spurious links artificially added are labeled with the 
"spurious_edge" tag using the code below. This labeling allows for tracking these links throughout
the analysis and network reconstruction process, although in typical use, such labeling would not
be necessary.

```{r add-spur}
for(i in 1:length(myInput$edgelists)){
    el <- myInput$edgelists[[i]]
    # Turn node ids into row and col indices
    rows <- as.numeric(el[, 1]) + 1
    cols <- as.numeric(el[, 2]) - nrow(myInput$data) + 1
    rows_cols <- paste0(rows, "_", cols)
    # Get spurious links indices
    spurious_rows_cols <- paste0(spurious_links[, 1], "_", spurious_links[, 2])
    spurious_loc <- which(rows_cols %in% spurious_rows_cols)
    # Add spurious edge type to edge lists
    el[spurious_loc, ]$x <- 1
    el[spurious_loc, ]$edge_type <- "spurious_edge"
    myInput$edgelists[[i]] <- el
}
```


```{r summary-input}
# Summarizes network characteristics
summary(myInput)
```


# Predict link probabilities with "marginal_all"

The `hsbm.predict` function applies the HSBM to predict marginal posterior probabilities of all 
links— both observed (`1s`) and unobserved (`0s`)—so you can identify spurious and missing links 
in one step. The function works directly with the processed input created by 
`hsbm.input`.Computation can be intensive, but parallelization across cores is supported.

```{r parallel-prediction}
# Generate HSBM predictions
myPred <- hsbm.predict(
  myInput,             # Input data processed by hsbm.input()
  iter = 10000,        # Number of iterations
  wait = 10000,         # Number of iterations for MCMC equilibration
  rnd_seed = 123,      # Sets seed in python environment for reproducibility
  method = "marginal_all", # Method for link prediction
  save_blocks = TRUE,  # Save group assignments 
  save_pickle = FALSE, # Save results as pickle files,
  save_plots = FALSE,  # Save hierarchical edge bundling plots
  n_cores = 3
)
```

```{r save-pred, include = FALSE}
save("myPred", file = "mypred_s3.RData")
```

Predicted link probabilities and group assignments are stored for each fold.

Below, we extract the link probabilities (*p*) for fold 1. For example, a high probability 
($p = 0.91$) of an undocumented link would suggest that this 
interaction is likely real but was missed due to sampling limitations. Conversely, it may 
assign a low *p* (e.g., 0.12) to a recorded interactions, indicating it 
might be spurious. 

```{r prediction-probs}
# View probabilities for fold 1
probabilities_fold1 <- myPred$probs[[1]]
head(probabilities_fold1)
```

The group assignments provide the hierarchical clustering structure of nodes for each fold. 
Let’s extract and examine the group assignments for fold 1:

```{r prediction-groups}
# View the group/block assignments for fold 1
groups_fold1 <- myPred$groups[[1]]

# Filter one type of nodes (nodes in columns)
vnames <- colnames(myPred$data)
groups_fold1 <- groups_fold1 %>% filter(name %in% vnames)
g_cols <- grep("^G", names(groups_fold1))
groups_fold1[g_cols] <- lapply(groups_fold1[g_cols], sort)

#print(head(groups_fold1))
print(lapply(myPred$groups[g_cols],
             function(x) apply(x, 2, function(y) length(unique(y)))))
```

Hierarchical group assignments provide insight into how nodes (e.g., hosts) are organized across 
multiple levels. At the first level (G1) nodes are divided into specific groups, reflecting
fine-scale patterns. Moving to higher levels (G2, G3, G4) these groups (or communities) are
progressively aggregated, revealing broader patterns and relationships or communities.

Next, we will explore how the model assigns specific probabilities *p* to different types of links
("documented," "held\_out," and "spurious\_edge") for each fold to evaluate HSBM ability to
correctly identify them. Documented links, which are real observations (true positives), should
have probabilities close to 1. Held-out links, which are real observations but transformed to 0s
during training for validation, should also show high probabilities. In contrast, spurious links,
artificially generated (initially unobserved links transformed to 1s), should have probabilities
close to 0 (unless it could be a missing link), as the model should recognize them as false links
(false negatives).

Since we use "marginal_all" method, all table numbers only take into account the links 
sampled during MCMC iterations. 

# Network Reconstruction

The `hsbm.reconstructed` function generates a reconstructed binary interaction matrix by combining predictions from all folds. The predicted matrix is transformed to binary values using a user-specified threshold.

```{r network-reconstruction, cache = FALSE}
# Network reconstruction
myReconst <- hsbm.reconstructed(
  myPred,                 # Predictions processed by hsbm.predict
  rm_documented=FALSE,    # Remove documented links during validation
  na_treatment = "ignore_na", # Handle NA values in predictions
  threshold = 0.5,            # Binarization threshold
  consistency_matrix = "average_thresholded" # Combine fold predictions
)
```

This output includes the final averaged probability matrix, the final reconstructed binary matrix, 
and evaluation metrics. These results highlight the predicted interactions, showcasing the
method's ability to detect spurious and missing links effectively. Let’s explore some of the key 
outputs.

```{r plot-reconstruction, cache = FALSE}
# Visualize the reconstructed binary matrix
plot_interaction_matrix(myReconst$new_mat, order_mat = FALSE)
print(myReconst$new_mat[sampled_indices])
```

```{r plot_reconstuction_spur, echo=TRUE}
# Plot the matrix with spurious
cols <- c("white", "red4")
dat_spur <- myReconst$data
new_spur <- which(myReconst$data == 1 & myReconst$new_mat == 0)
new_missing <- which(myReconst$data == 0 & myReconst$new_mat == 1)
if(length(new_spur) != 0){
    dat_spur[new_spur] <- 2
    cols <- c(cols, "blue4")
}
if(length(new_missing)!=0){
    dat_spur[new_missing] <- 3
    cols <- c(cols, "green4")
}
plot_interaction_matrix(dat_spur,
                        col = cols, order_mat = FALSE)
```


```{r heatmap}
library(ggplot2)
library(RColorBrewer)

mat_avg <- apply(simplify2array(myReconst$pred_mats), 1:2, mean,
                 na.rm = TRUE)
print(mat_avg[sampled_indices])

melted_matrix <- reshape2::melt(mat_avg)

row_names <- rownames(mat_avg)
colnames(melted_matrix) <- c("Rows", "Cols", "Probs")
ggplot(melted_matrix, aes(Cols, Rows, fill= Probs)) + 
  geom_tile() + 
  scale_y_discrete(limits = rev(row_names)) + 
  scale_fill_distiller(palette = "Spectral", direction = 1) +
  theme_minimal() +
  theme(axis.text = element_blank())
```

```{r summary-reconstruction, cache = FALSE}
# View the reconstruction summary and evaluation metrics0
summary(myReconst)
```

The summary provides the number of spurious and missing links, as well as key evaluation metrics, 
such as the recovery link ratio (RLF).

Let's now examine the top 10 predicted links that are most likely to be spurious (false positives)
by visualizing the probabilities of "documented" links.

```{r top-links_spur, cache =  FALSE}
# Visualize the top most likely spurious links
top_links_spurious <- top_links(myReconst, 
                                n = 10, 
                                edge_type = "documented")
print(top_links_spurious)
```

This example was tailored to find spurious links and the model didn't identify any missing links
at the threshold of 0.5, which is a very conservative threshold. However we note that on 
average, `r round(mean(count_higher/res_summary[, 8])*100, 0)`% of the held-out 
links had a higher probability than all undocumented links across folds.

We show how we can rank the most likely missing links according to the model.

```{r top-links_missing, cache =  FALSE}
# Visualize the top most likely spurious links
top_links_missing <- top_links(myReconst, 
                               n = 10, 
                               edge_type = "undocumented")
print(top_links_missing)
```

If the task is to do network reconstruction by inserting the most plausible missing links, 
we can use the probabilities from the `marginal_all` method as 'scores' in a 
supervised binary classification, which is implemented only for missing links. 
This is achieved by following the steps in Supporting Information S1, but setting 
"method = 'marginal\_all'" in `hsbm.predict()` and optionally "rm\_documented = 'FALSE'" in 
`hsbm.reconstructed()`.

This document demonstrates the use of the *sabinaHSBM* package for network reconstruction. 
By applying the "marginal\_all" method, we showcased how likely spurious and missing links can be
identified and addressed in complex networks.

# Testing....

```{r network-reconstruction2}
# Network reconstruction
myReconst <- hsbm.reconstructed(
  myPred,                # Predictions processed by hsbm.predict
  rm_documented = FALSE,  # Use of documented entries during validation
)
print(myReconst$stats)
```

# Computing characteristics

```{r end_time}
# Show processing time and computer characteristics (optional)
end_time <- Sys.time()
cat("The processing time of this script took: ",
    round(as.numeric(difftime(end_time, start_time, units = "mins", 1))), "minutes\n")
```

This analysis was performed on a Dell Inspiron notebook with the following
characteristics:

- Processor (CPU): Intel Core i7-7500U @ 3.5GHz 
- Memory (RAM): 16 GB
- Operating System: Ubuntu 24.04 (run on a docker image)
- R Version: 4.3.3
